---
title: "rgi-viz"
output: html_document
params:
  mapq_thresh: 100
  coverage_thresh: 75
  numreads_thresh: 5000
---

```{r}
colors = c("#8a2f2f",
           "#c5ae93", "#eda8b7", "#828cb4", "#45aad5", "#26576f", "#5468ba", "#e06781", "#a98760", "#36837e")

colors[2:10]
```


```{r}
pacman::p_load(dplyr, ggplot2, ComplexHeatmap, vegan, usedist, tidyr, circlize)

gmd <- read.delim("all-stations.gene_mapping_data.txt")

meta <- read.csv("data/clean_metadata.csv") %>% 
  filter(uuid %in% unique(gmd$id)) %>% 
  tibble::column_to_rownames(var = "uuid")

mechanisms <- c("antibiotic target alteration" = "зміна мішені", "antibiotic inactivation" = "інактивація", "antibiotic efflux" = "виведення", "antibiotic target alteration; antibiotic target replacement" = "зміна + заміна мішені", "reduced permeability to antibiotic" = "зниження проникності", "antibiotic target protection" = "захист мішені", "antibiotic target replacement" = "заміна мішені", "antibiotic efflux; reduced permeability to antibiotic" = "виведення + зн.проник.", "antibiotic target alteration; antibiotic efflux" = "зміна мішені + виведення")

arometa <- gmd %>%
  select(ARO.Term, Resistance.Mechanism, AMR.Gene.Family, Drug.Class) %>% 
  filter(!duplicated(.)) %>% arrange(ARO.Term) %>% 
  tibble::column_to_rownames(var = "ARO.Term") %>% 
  mutate("Механізм резистентності" = mechanisms[arometa$Resistance.Mechanism])

abund <- gmd %>%
  select(ARO.Term, Completely.Mapped.Reads, id) %>%
  tidyr::pivot_wider(names_from = "id", values_from = Completely.Mapped.Reads) %>% 
  #filter(ARO.Term %in% aroterms) %>% 
  tibble::column_to_rownames(var = "ARO.Term")

abund %>% head(2)

```

# Coverage VS Mapped reads plot

```{r fig.width=6, fig.height=6}
mapq_thresh = 100
coverage_thresh = 75
numreads_thresh = 5000

gmd %>%
  filter(Average.MAPQ..Completely.Mapped.Reads. > mapq_thresh) %>%
  mutate(
    thresh_group = case_when(
      All.Mapped.Reads > numreads_thresh & Average.Percent.Coverage > coverage_thresh ~ "both",
      All.Mapped.Reads > numreads_thresh ~ "numreads_pass",
      Average.Percent.Coverage > coverage_thresh ~ "coverage_pass",
      TRUE ~ "fail")
    ) %>%
  ggplot(aes(x = All.Mapped.Reads,
             y = Average.Percent.Coverage,
             color = thresh_group)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_vline(xintercept = numreads_thresh, linetype = 1, color = "gray30") +
  geom_hline(yintercept = coverage_thresh, linetype = 1, color = "gray30") +
  scale_color_manual(values = c("both" = "#FF7F0F", 
                                "numreads_pass" = "#7681B3", 
                                "coverage_pass" = "#AD9F71", 
                                "fail" = "gray80")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Кількість прочитань",
       y = "Покриття гену, %",
       title = "")

```

```{r}
# a list of AR terms that passed the filter for at least a single sample
aroterms <- gmd %>% filter(Average.MAPQ..Completely.Mapped.Reads. > params$mapq_thresh &
               (Completely.Mapped.Reads > params$numreads_thresh |
               Average.Percent.Coverage > params$coverage_thresh)) %>% 
  pull(ARO.Term)
```

# Permanova

```{r}
nreads_df <- gmd %>% 
  filter(ARO.Term %in% aroterms) %>% 
  select(id, ARO.Term, Completely.Mapped.Reads) %>% 
  tidyr::pivot_wider(names_from = ARO.Term, values_from = Completely.Mapped.Reads) %>% 
  tibble::column_to_rownames(var = "id")
head(nreads_df, 2)
```

## Between stations

```{r}
# Bray-Curtis (abundance-based)
bray_dist <- vegdist(nreads_df, method = "bray", na.rm = TRUE)
# Permanova on Bray-Curtis dissimilarity matrix
bstation <- adonis2(bray_dist ~ station, data = meta, permutations = 999)
bstation
```

```{r}
# Jaccard (presence/absence)
binary_matrix <- as.data.frame((nreads_df > 0) * 1)
binary_matrix[is.na(binary_matrix)] <- 0
jaccard_dist <- vegdist(binary_matrix, method = "jaccard", na.rm = TRUE)

jstation <- adonis2(jaccard_dist ~ station, data = meta, permutations = 999)
jstation
```

# Surface

```{r}
table(meta$surface)
```

Only the surfaces with at least 10 samples are tested.

```{r}
surface_subs <- meta %>%
  filter(surface %in% c("bench", "entrance_door", "handrail", "ticket_machine")) %>% 
  rownames()

bsurface <- adonis2(dist_subset(bray_dist, surface_subs) ~ surface,
        data = meta[surface_subs, ], permutations = 999)
bsurface
```

```{r}
jsurface <- adonis2(dist_subset(jaccard_dist, surface_subs) ~ surface,
        data = meta[surface_subs, ], permutations = 999)
jsurface
```

## Material

```{r}
material_subs <- meta %>%
  filter(surface_material %in% c("metal", "plastic", "wood")) %>% 
  rownames()

bmaterial <- adonis2(dist_subset(bray_dist, material_subs) ~ surface_material,
        data = meta[material_subs, ], permutations = 999)

bmaterial
```

!!!

```{r}
jmaterial <- adonis2(dist_subset(jaccard_dist, material_subs) ~ surface_material,
        data = meta[material_subs, ], permutations = 999)
jmaterial
```

```{r}
sink("out/AMR-analysis/PERMANOVA_AMR-profiles.txt")
bstation
jstation
bsurface
jsurface
bmaterial
jmaterial
sink()
```

## Visualization of a significant result

```{r fig.width= 8, fig.height = 8}
jmat <- as.matrix(dist_subset(jaccard_dist, material_subs))

split <- meta[rownames(jmat), "Матеріал"]
split_colors <- setNames(rainbow(length(unique(split))), unique(stations))

Heatmap(jmat,
        name = "Індекс\nЖаккара",
        show_column_names = F, show_row_names = F,
        cluster_rows = T, cluster_columns = T,
        split = split)
```

# Basic stats

```{r}
arostat <- abund %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%  
  mutate(
    abund = rowSums(., na.rm = T),
    prev = rowSums(. > 0, na.rm = T),
    prevperc = rowMeans(. > 0, na.rm = T)*100) %>%
  merge(arometa, by = "row.names") %>% 
  select(!starts_with("haib"))

arostat %>% group_by(Resistance.Mechanism) %>%
  summarize(abund = sum(abund),
            aroterms = n_distinct(Row.names)) %>%
  arrange(desc(abund))

arrange(arostat, desc(abund)) %>% head(20)

filter(arostat, prevperc == 100) %>% arrange(desc(abund))

```
```{r}
# save core AMR table
filter(arostat, prevperc == 100) %>% arrange(desc(abund)) %>% 
  write.csv("out/AMR-analysis/core-AMR-genes.csv")
```

```{r}
gmd %>% filter(ARO.Term == "qacG")
gmd %>% filter(Resistance.Mechanism == "antibiotic target alteration; antibiotic target replacement") %>% 
  pull(ARO.Term) %>% unique()
```

```{r}

  #select(species, abund, prevperc) %>% 
  
prevstat %>%   
group_by(Resistance.Mechanism) %>% 
  summarize(abund = sum(abund)) %>% 
  arrange(desc(abund)) %>% 
  rename("Механізм резистеності" = Resistance.Mechanism,
         "Загальна кількість прочитань" = abund)
  
prevstat %>% 
  arrange(desc(abund))
```

```{r}
summarize_amr <- function(group_var) {
  gmd %>%
    select(ARO.Term, Completely.Mapped.Reads, id) %>%
    tidyr::pivot_wider(names_from = "ARO.Term", values_from = Completely.Mapped.Reads) %>%
    tibble::column_to_rownames(var = "id") %>%
    mutate(
      across(everything(), ~tidyr::replace_na(., 0)),
      !!group_var := meta[rownames(.), group_var]  # dynamically insert column name
    ) %>%
    group_by(.data[[group_var]]) %>%   # group by the dynamic column
    summarize(across(where(is.numeric), sum), .groups = "drop") %>%
    mutate(
      "Прочитань" = rowSums(select(., where(is.numeric)), na.rm = TRUE),
      "АМР генів" = rowSums(select(., where(is.numeric)) > 0, na.rm = TRUE)
    ) %>%
    select(all_of(group_var), "Прочитань", "АМР генів") %>%
    arrange(desc(`АМР генів`))
}
summarize_amr("Станція")
summarize_amr("Матеріал")
summarize_amr("Поверхня")

```

## kruskal wallis between materials by number of AMR genes

```{r}
arocount_material <- gmd %>%
  select(ARO.Term, Completely.Mapped.Reads, id) %>%
  pivot_wider(names_from = "ARO.Term", values_from = Completely.Mapped.Reads) %>%
  column_to_rownames(var = "id") %>%
  mutate(arocount = rowSums(. > 0, na.rm = T),
         material = meta[rownames(.), "Матеріал"]) %>% 
  select(arocount, material)
head(arocount_material, 2)
```

```{r}
arocount_material %>% 
  filter(material %in% c("дерево", "метал", "пластик")) %>% 
  group_by(material) %>%
  summarise(shapiro_test = shapiro.test(arocount)$p.value)
```

```{r}
kruskal.test(arocount ~ material, data = arocount_material)
```


## Adundance of selected amr genes

```{r}
abundmat <- abund %>% filter(rownames(.) %in% aroterms) %>% data.matrix %>% log10()
#abundmat[is.na(abundmat)] <- min(abundmat, na.rm = TRUE) - 1

colannot <- HeatmapAnnotation(which = "column",
    "Матеріал" = meta[colnames(abundmat), "Матеріал"],
    "Станції" = meta[colnames(abundmat), "Станція"])

```

```{r fig.width=7, fig.height=6}
mech <- factor(arometa[rownames(abundmat), "Механізм резистентності"])
mech_colors <- setNames(colors[2:7], levels(mech))
rowannot <- rowAnnotation(
   show_legend = FALSE,
   Механізм = mech,
   col = list(Механізм = mech_colors),
  annotation_name_gp = gpar(fontsize = 0),
  annotation_legend_param = list(
    Механізм = list(ncol = 2, break_dist = 880)
  )
)


col_fun = colorRamp2(c(0, 6), c("white", "#8A2F2F"))


hm <- Heatmap(abundmat,
              col = col_fun,
              row_labels = abbreviate(rownames(abundmat), minlength = 10),
              cluster_rows = F, cluster_columns = F,
              name = "Прочитань, log10",
              row_names_gp = gpar(fontsize = 8),
        right_annotation = rowannot,
        show_row_names = T,  show_column_names = F,
        heatmap_legend_param = list(
          direction = "vertical",
          #col_fun = col_fun,
          title = "Прочитань, log10",
          title_position = "lefttop-rot",
          legend_height = unit(5, "cm")))


mech_legend <- Legend(
  title = "Механізм",
  at = names(mech_colors),
  labels = names(mech_colors),
  legend_gp = gpar(fill = mech_colors),
  ncol = 3,
  gap = unit(20, "mm"),  # Controls spacing between columns
  labels_gp = gpar(fontsize = 10)
)

draw(hm,
     heatmap_legend_side = "left",
     annotation_legend_list = list(mech_legend),
     annotation_legend_side = "bottom")
```

```{r}
cairo_pdf(file = "figures/amr-heatmap.pdf",
width = 7, # The width of the plot in inches
    height = 6) # The height of the plot in inches
draw(hm,
     heatmap_legend_side = "left",
     annotation_legend_list = list(mech_legend),
     annotation_legend_side = "bottom")
dev.off()
```

## Box plots of reads attributed to AMR genes / number of AMR genes per mechanism

```{r fig.width=3.5, fig.height=3.5}
b1 <- abund %>% 
  tibble::rownames_to_column(var = "aro") %>% 
  pivot_longer(!aro, names_to = "samples", values_to = "abund") %>% 
  mutate("Механізм" = arometa[.$aro, "Механізм резистентності"]) %>% 
  ggplot(aes(x = Механізм, y = abund, fill = Механізм)) +
  geom_jitter(aes(color = Механізм),
              width = 0.2, alpha = 0.7, size = 1, show.legend = FALSE) +
  geom_boxplot(outlier.shape = NA, alpha = 0) +
  #scale_y_continuous(name = "Абунданс") +
  #scale_x_discrete(name = "Механізм резистентності") +
  scale_y_log10() +
  labs(x = NULL, y = "Прочитань, log10") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
   coord_flip() +
  scale_color_manual(values = colors[2:10])

b1
```

```{r fig.width=3.5, fig.height=3.5}
b2 <- abund %>% mutate(nsamples = rowSums(. > 0, na.rm = T)) %>% 
  select(nsamples) %>% 
  mutate("Механізм" = arometa[rownames(.), "Механізм резистентності"]) %>% 
  ggplot(aes(x = Механізм, y = nsamples, fill = Механізм)) +
  geom_jitter(aes(color = Механізм),
              width = 0.2, alpha = 0.7, size = 1, show.legend = FALSE) +
  geom_boxplot(outlier.shape = NA, alpha = 0) +
  #scale_y_continuous(name = "Абунданс") +
  #scale_x_discrete(name = "Механізм резистентності") +
  labs(x = NULL, y = "Зразків") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
   coord_flip() +
  scale_color_manual(values = colors[2:10])

b2
```


```{r fig.width=4, fig.height=3.5}
library(patchwork)

b1 + b2 + plot_layout(axes = "collect")
```

```{r}
cairo_pdf(file = "figures/amr-boxplots.pdf",
width = 4, # The width of the plot in inches
    height = 3.5)
b1 + b2 + plot_layout(axes = "collect")
dev.off()
```

# Heatmap AMR x Sample coverage %

```{r coverage matrix}
covmat <- gmd %>% filter(ARO.Term %in% aroterms) %>% select(ARO.Term, id, Average.Percent.Coverage) %>%
  tidyr::pivot_wider(names_from = ARO.Term, values_from = Average.Percent.Coverage) %>% 
  tibble::column_to_rownames(var = "id") %>%
  data.matrix()
```

```{r row and col annots}
row_annot <- HeatmapAnnotation(which = "row",
    "Станція" = meta[match(rownames(covmat), meta$uuid), "station"])

# annotation df for amr genes
aroannot_df <- gmd %>%
  distinct(ARO.Term, .keep_all = TRUE) %>%
  filter(ARO.Term %in% colnames(covmat))

col_annot <- HeatmapAnnotation(which = "column",
    "Механізм резистентності" = aroannot_df[match(colnames(covmat), aroannot_df$ARO.Term), "Resistance.Mechanism"])
```

```{r}
Heatmap(covmat, cluster_rows = F, cluster_columns = F, name = "Покриття, %",
        right_annotation = row_annot, top_annotation = col_annot,
        show_row_names = F)

#draw(hm, column_title = "Кількість прочитань")
```
- [] renames resistance mechanism
- [] rename stations
- [] move annotation labels from legend to the bar
- [] order by col annot
- [] add barplot to columns (number of AMRs identified)
- [] recolor


```{r}
# Match stations to rownames in covmat
stations <- meta[match(rownames(covmat), meta$uuid), "station"]
station_colors <- setNames(rainbow(length(unique(stations))), unique(stations))

mechs <- aroannot_df[match(colnames(covmat), aroannot_df$ARO.Term), "Resistance.Mechanism"]
mech_colors <- setNames(rainbow(length(unique(mechs))), unique(mechs))
  
  
row_annot <- rowAnnotation(
  Станція = anno_block(
    gp = gpar(fill = station_colors),
    labels = names(station_colors),
    labels_gp = gpar(fontsize = 10),
    labels_rot = 90,
    width = unit(5, "mm")
  )
)

col_annot <- HeatmapAnnotation(
  "Механізм резистентності" = anno_block(
    gp = gpar(fill = mech_colors),
    labels = names(mech_colors),
    labels_gp = gpar(fontsize = 10),
    labels_rot = 0,
    width = unit(5, "mm")
  )
)

# Final heatmap with row split
Heatmap(covmat,
        cluster_rows = FALSE, cluster_columns = FALSE,
        name = "Покриття, %",
        split = stations,
        column_split = mechs,
        right_annotation = row_annot,
        top_annotation = col_annot,
        show_row_names = FALSE,
        row_title = NULL, column_title = NULL)

```

