
- [x] bracken pca
- [x] barplot family stacked
- [x] Jaccard dist heatmap
- [x] unclassified distrib

```{r}
pacman::p_load("dplyr", "tidyr", "ggplot2", "factoextra", "stringr", "forcats",
               "janitor", "ggfortify", "pheatmap", "vegan", "patchwork",
               "ComplexHeatmap")

bracken <- read.delim("out/taxonomy-analysis/combined_out.bracken") %>%
  select(taxonomy_id, ends_with("filtered_num")) %>% 
  rename_with(~ str_remove_all(.x, "haib17CEM5241_HMGMHCCXY_|\\.out\\.bracken\\.filtered_num$"))

taxt <- read.delim("out/DA_analysis/taxtable.tsv", header = F,
                   col.names = c("taxonomy_id",  "Kingdom", "Phylum", "Class", "Order",
                                 "Family", "Genus", "Species", "Subspecies"))

bracken <- left_join(bracken, taxt, by="taxonomy_id")
```

# Bracken PCA

```{r}
meta <- read.csv("data/complete_metadata.csv") %>% 
  filter(city == "kyiv",
         !(station %in% c("busstop_20", ""))) %>% 
  select_if(sapply(., function(x) !all(is.na(x)), USE.NAMES = F)) %>% 
  mutate(id = word(uuid, 3, sep = "_"),
         idcopy = id,
         station = gsub("petrivka", "Pochaina", station),
         station = stringr::str_to_title(gsub("_", " ", station))) %>% 
  filter(!(id %in% id[duplicated(id)])) %>% 
  tibble::column_to_rownames(var = "idcopy")
```

```{r}
pcadata <- bracken %>% 
  select(taxonomy_id, starts_with("SL"))%>% t() %>%
  as.data.frame() %>% row_to_names(1)

pca <- prcomp(pcadata, scale. = T)
pcadata
```
```{r}
pcadata_annot <- pcadata %>%
  tibble::rownames_to_column(var = "id") %>% 
  left_join(meta, by = "id")
```

```{r}
fviz_eig(pca)
```

```{r}
autoplot(pca, data = pcadata_annot, colour = "surface_material")
autoplot(pca, data = pcadata_annot, colour = "surface")
autoplot(pca, data = pcadata_annot, colour = "station")

```

# PCA on RPM normalized

```{r}
pcadata_norm <- pcadata %>%
  mutate(across(everything(), ~.x/sum(.x) * 1000000))

pca_norm <- prcomp(pcadata_norm)

head(pcadata_norm)
```

```{r}
fviz_eig(pca_norm)
```

```{r}
autoplot(pca_norm, data = pcadata_annot, colour = "surface_material")
autoplot(pca_norm, data = pcadata_annot, colour = "surface")
autoplot(pca_norm, data = pcadata_annot, colour = "station")

```

The outlier on the PCA plot

```{r}
pca_norm$x %>% as.data.frame() %>% 
  filter(PC1 > 25*10^6) %>%
  merge(meta, by="row.names") %>% 
  select(!starts_with("PC"))
```

```{r}
hist(data.frame(pca_norm$x)$PC1)
```


```{r}
pca_norm_filt <- 
pca_norm$x %>% as.data.frame() %>% 
  filter(PC1 < 25*10^6) %>% 
  merge(meta, by="row.names")

ggplot(data = pca_norm_filt, aes(x=PC1, y=PC2, colour = factor(station))) +
  geom_point()
```

# Bracken out stacked bar

## All samples

```{r}
bracken %>% colnames()
```


## By station

```{r}
famsums <- bracken %>% group_by(Family) %>% 
  summarize(across(
    .cols = starts_with("SL"), 
    .fns = sum, 
    .names = "{.col}"
    )) %>% 
  mutate(totabund = rowSums(select_if(., where(is.numeric)))) %>% 
  filter(totabund > 0) %>% 
  mutate(Family = if_else(totabund / sum(totabund) > 0.01, Family, "Other")) %>% 
  group_by(Family) %>% 
  summarize(across(
    .cols = is.numeric, 
    .fns = sum))

head(famsums, 3)
```

```{r}
famsums_long <- famsums %>% 
  pivot_longer(!Family, names_to = "sample", values_to = "count") %>%
  mutate(Family = fct_reorder(Family, count, .fun = sum, .desc = TRUE)) %>% 
  left_join(meta %>% tibble::rownames_to_column(var = ".sample"), by = c("sample" = ".sample")) %>% 
  mutate(sample = if_else(sample == "totabund", "", sample),
         station = if_else(is.na(station), "Total Abundance", station))

ggplot(famsums_long, aes(fill = Family, y = count, x = sample)) + 
    geom_bar(position = "fill", stat = "identity") +
    labs(x = NULL, y = NULL) +
    theme_void() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),  # Rotate x-axis labels
        legend.title = element_blank()  # Remove legend title for a cleaner look
    ) + facet_wrap(vars(station), scales = "free")
```

# The disctribution of unclassified reads

```{r}
uncl_reads <- read.delim("out/taxonomy-analysis/multiqc_kraken/multiqc_data/kraken-top-n-plot_Species.txt") %>% 
  select(Unclassified)

ggplot(uncl_reads, aes(x=Unclassified)) + 
  geom_histogram(bins = 15)
```
# Hist number of reads

```{r}
ggplot(meta, aes(x=num_reads)) + 
  geom_histogram(bins = 15)
```

```{r}
bardata <- 
read.delim("out/taxonomy-analysis/multiqc_kraken/multiqc_data/kraken-top-n-plot_Species.txt",
           check.names = F) %>% 
  mutate(id = word(Sample, 3, sep = "_")) %>% 
  select(!Sample) %>% 
  mutate(Classified = rowSums(select(., -c(id, Unclassified)))) %>% 
  select(id, Classified, Unclassified) %>% 
  rename("")
  mutate(id = fct_reorder(id, Classified))  %>%
  pivot_longer(!id) %>%
  mutate(name = factor(name, levels = c("Unclassified", "Classified"))) 
```

```{r fig.height=8, fig.width=9}
p1 <- ggplot(bardata, aes(fill=name, y=id, x=value)) + 
    geom_bar(position="stack", stat="identity") +
  labs(title = "Кількість прочитань", y = NULL, x = NULL) +
  theme_minimal()

p2 <- percreads <- ggplot(bardata, aes(fill=name, y=id, x=value)) + 
    geom_bar(position="fill", stat="identity") +
  labs(title = "Пропорція прочитань", y = NULL, x =NULL) +
  theme_minimal()

p1 + p2 + 
  plot_layout(guides = "collect", axes = "collect_y") &
  theme(legend.position = "bottom")
```

# Jaccard distance heatmap on Bracken out

```{r}
bracken <- 
read.delim("out/taxonomy-analysis/combined_out.bracken") %>%
  tibble::column_to_rownames(var="taxonomy_id") %>% 
  select(ends_with("filtered_num")) %>% 
  rename_with(~ str_remove_all(.x, "haib17CEM5241_HMGMHCCXY_|\\.out\\.bracken\\.filtered_num$"))

samples_order <- colnames(bracken)

distm <- t(bracken) %>% vegdist(method = "jaccard", diag = T, upper = T) %>%
  as.matrix()
colnames(distm) <- samples_order
rownames(distm) <- samples_order
```

```{r}
annotdf <- read.delim("out/taxonomy-analysis/multiqc_kraken/multiqc_data/kraken-top-n-plot_Species.txt",
           check.names = F) %>%
  mutate(tot = rowSums(select(., -Sample))) %>% 
  mutate(id = word(Sample, 3, sep = "_")) %>% 
  select("id", "Homo sapiens", "Unclassified", "tot") %>% 
  left_join(meta) %>% 
  tibble::column_to_rownames(var="id") %>% 
  mutate("Human reads, %" = .$"Homo sapiens" / tot,
         "Human reads, log" = log10(.$"Homo sapiens"),
         "Unclassified reads, %" = .$"Unclassified" / tot)

annotdf
```
```{r fig.height=12, fig.width=12}
ha = HeatmapAnnotation("Total reads" = annotdf[samples_order, "tot"],
                       "Unclassified reads" = annotdf[samples_order, "Unclassified"],
                       "Unclassified reads, %" = annotdf[samples_order, "Unclassified reads, %"],
                       "Human reads" = annotdf[samples_order, "Homo sapiens"],
                       "Human reads, %" = annotdf[samples_order, "Human reads, %"],
                       "Human reads, log" = annotdf[samples_order, "Human reads, log"],
                       "Flowcell" = annotdf[samples_order, "hudson_alpha_flowcell"])
Heatmap(distm, name = "Jaccard dist", top_annotation = ha)
```

```{r fig.height=12, fig.width=12}
ha = HeatmapAnnotation("Total reads" = annotdf[samples_order, "tot"],
                       "Unclassified reads" = annotdf[samples_order, "Unclassified"],
                       "Unclassified reads, %" = annotdf[samples_order, "Unclassified reads, %"],
                       "Human reads" = annotdf[samples_order, "Homo sapiens"],
                       "Human reads, %" = annotdf[samples_order, "Human reads, %"],
                       "Human reads, log" = annotdf[samples_order, "Human reads, log"],
                       "Flowcell" = annotdf[samples_order, "hudson_alpha_flowcell"])
hm <- Heatmap(distm, name = "Jaccard dist", top_annotation = ha, column_km = 3, row_km = 3)
hm <- draw(hm)
```
```{r}
# get the clusters for Ancom
clustl = column_order(hm)
clustdf <- lapply(names(clustl), function(i){
  out <- data.frame(id = colnames(distm)[clustl[[i]]],
                    clust = i, stringsAsFactors = FALSE)
     return(out)
   }) %>% 
  do.call(rbind, .)

write.table(clustdf, file = "out/DA_analysis/clust_complete_euclidean.tsv", sep = "\t", row.names = F) 
```

```{r}
meta$hudson_alpha_flowcell %>% unique()
```
## Without human reads

```{r fig.height=12, fig.width=12}
bracken_nh <- subset(bracken, rownames(bracken) != 9606)
samples_order_nh <- colnames(bracken_nh)

distm_nh <- bracken_nh %>% t()%>%
  vegdist(method = "jaccard", diag = T, upper = T) %>% as.matrix()
colnames(distm_nh) <- samples_order_nh
rownames(distm_nh) <- samples_order_nh

ha_nh = HeatmapAnnotation("Total reads" = annotdf[samples_order_nh, "tot"],
                       "Unclassified reads" = annotdf[samples_order_nh, "Unclassified"],
                       "Unclassified reads, %" = annotdf[samples_order_nh, "Unclassified reads, %"],
                       "Human reads" = annotdf[samples_order_nh, "Homo sapiens"],
                       "Human reads, %" = annotdf[samples_order_nh, "Human reads, %"],
                       "Human reads, log" = annotdf[samples_order_nh, "Human reads, log"],
                       "station" = annotdf[samples_order_nh, "station"],
                       "surface material"= annotdf[samples_order_nh, "surface_material"])

Heatmap(distm_nh, name = "Jaccard dist", top_annotation = ha_nh)
```

```{r}
meta <- read.csv("data/complete_metadata.csv") %>% filter(city == "kyiv")
colnames(meta)
```

```{r}
meta[, unlist(lapply(meta, function(x) !all(is.na(x))))]
```

```{r}
table(meta$hudson_alpha_flowcell)

meta %>% 
  select_if(sapply(., function(x) !all(is.na(x)), USE.NAMES = F))
```

# Heatmap of counts per sample

```{r}
summary(rowSums(bracken))
```

```{r}
brackmat <- bracken_nh %>% arrange(desc(rowSums(.))) %>% head(200) %>% as.matrix()

# calculate proportions of reads attributed to taxon
brackmat <- sweep(brackmat, 2, colSums(brackmat), FUN = "/")
```

```{r fig.height=12, fig.width=12}
ha = HeatmapAnnotation("Станція" = annotdf[samples_order, "station"],
                       "Кількість прочитань" = annotdf[samples_order, "num_reads"],
                       "Поверхня" = annotdf[samples_order, "surface"],
                       "Матеріал" = annotdf[samples_order, "surface_material"],
                       "Homo Sapiens прочитань, %" = annotdf[samples_order, "Human reads, %"])
hm <- Heatmap(brackmat, name = "Прочитань", top_annotation = ha,
        show_row_names = F, show_row_dend = F)

draw(hm, column_title = "Кількість прочитань", cluster)
```

# Heatmap of correlation between samples

```{r}
#install.packages('pcaPP')
cormat <- pcaPP::cor.fk(bracken)
```

```{r fig.height=12, fig.width=12}
ha = HeatmapAnnotation("Станція" = annotdf[samples_order, "station"],
                       "Поверхня" = annotdf[samples_order, "surface"],
                       "Матеріал" = annotdf[samples_order, "surface_material"],
                       "Загальна кількість прочитань" = annotdf[samples_order, "num_reads"],
                       "Homo Sapiens прочитань, %" = annotdf[samples_order, "Human reads, %"])
hm <- Heatmap(cormat, name = "Коеф.\nкор.", top_annotation = ha,
        show_row_names = F, show_row_dend = F)

draw(hm, column_title = "Кореляція Кендала")
```

# Heatmap of Jaccard distance between sample prevalence profiles

```{r}
bi_brackmat <- ifelse(as.matrix(bracken) == 0, 0, 1)

distm_bi <- bi_brackmat %>% t()%>%
  vegdist(method = "jaccard", diag = T, upper = T) %>% as.matrix()
colnames(distm_bi) <- samples_order
rownames(distm_bi) <- samples_order
```

```{r fig.height=12, fig.width=12}
hm <- Heatmap(distm_bi, name = "Відстань", top_annotation = ha,
              show_row_names = F, show_row_dend = F)
draw(hm, column_title = "Відстань Жаккара між бінарними профілями присутності зразків")
```

# Taxonomy stats

```{r}
library(dplyr)
bracken %>% left_join(ktax, by = rownames())

brackannot <- bracken %>% merge(ktax, by=0)

sstats <- brackannot %>%
  mutate(tot_counts = rowSums(select(., starts_with("SL")))) %>% 
  rowwise() %>%
  mutate(num_samples = sum(c_across(where(is.numeric)) > 0)) %>% 
  select(Species, Genus, tot_counts, num_samples) %>% 
  arrange(desc(tot_counts))

hist(sstats$tot_counts)
summary(sstats$tot_counts)

sstats %>% filter(tot_counts > quantile(.$tot_counts, .995))
```
Набір видів, що є спільними для всіх 69(?) зразків, складається з 52 видів. Серед них 
```{r}
brackannot %>%
  mutate(tot_counts = rowSums(select(., starts_with("SL")))) %>% 
  rowwise() %>%
  mutate(num_samples = sum(c_across(where(is.numeric)) > 0)) %>% 
  select(Species, Genus, tot_counts, num_samples) %>% 
  filter(num_samples == max(.$num_samples))
```

Most prevalent genus

```{r}
sstats %>% group_by(Genus) %>% summarise(tot_counts = sum(tot_counts)) %>% 
  arrange(desc(tot_counts)) %>% filter(tot_counts > quantile(.$tot_counts, 0.99))
```

