---
params:
  classif_lvl: "G"
  abundtable: "out/taxonomy-analysis/combined_out.bracken"
  metadata: "data/clean_metadata.csv"
  taxtable: "out/DA_analysis/taxtable.tsv"
  rank: "Genus"
---

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#update.packages("BiocManager")
#BiocManager::install("ANCOMBC", lib="~/R/x86_64-pc-linux-gnu-library/4.4")

#BiocManager::install("microbiome")
```

```{r}
library(dplyr)
library(phyloseq)
library(tibble)
library(ANCOMBC)
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(tibble)
library(vegan)
library(Cairo)
library(usedist)
library(stringr)

# extract taxid identified to filter taxonomy table
taxids <- read.delim(params$abundtable) %>% pull(taxonomy_id)

abund_df <- read.delim(params$abundtable) %>%
  column_to_rownames("taxonomy_id") %>% 
  select(ends_with("filtered_num")) %>% 
  rename_with(~ gsub(".out.bracken.filtered_num", "", .x))

meta <- read.csv(params$metadata) %>% 
  filter(uuid %in% colnames(abund_df)) %>% 
  column_to_rownames("uuid")

ktax_cols <- c("taxid", "root", "domain", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Subspecies")
ktax <- read.delim(params$taxtable, header = F,
                   col.names = ktax_cols) %>% 
  column_to_rownames("taxid")
```
# Sample categories table

```{r}
library(rlang)
for(group in c("Станція", "Поверхня", "Матеріал")){
  meta %>% filter(surface != "floor") %>% group_by(!!sym(group)) %>%
    summarize("Зразків" = n()) %>% arrange(desc(Зразків)) %>% print()
}
```

# Permanova

```{r}
nreads_df <- 
abund_df %>%
  tibble::rownames_to_column(var = "taxid") %>% 
  tidyr::pivot_longer(!taxid) %>%
  tidyr::pivot_wider(names_from = taxid, values_from = value) %>%
  tibble::column_to_rownames(var = "name")
head(nreads_df, 2)
```
## Between stations

```{r}
# Bray-Curtis (abundance-based)
bray_dist <- vegdist(nreads_df, method = "bray", na.rm = TRUE)
# Permanova on Bray-Curtis dissimilarity matrix
adonis2(bray_dist ~ station, data = meta, permutations = 999)
```

```{r}
# Jaccard (presence/absence)
binary_matrix <- as.data.frame((nreads_df > 0) * 1)
binary_matrix[is.na(binary_matrix)] <- 0
jaccard_dist <- vegdist(binary_matrix, method = "jaccard", na.rm = TRUE)

adonis2(jaccard_dist ~ station, data = meta, permutations = 999)
```

W/o car samples

```{r}
station_samples <- meta %>% filter(station_specific != "car") %>% rownames()
adonis2(dist_subset(bray_dist, station_samples) ~ station,
        data = meta[station_samples, ], permutations = 999)
```

```{r}
adonis2(dist_subset(jaccard_dist, station_samples) ~ station,
        data = meta[station_samples, ], permutations = 999)
```

## Surfaces

```{r}
meta %>%
  filter(#surface %in% c("bench", "entrance_door", "handrail", "ticket_machine"),
         station_specific == "car")
```

Only the surfaces with at least 11 samples are tested.

```{r}
surface_subs <- meta %>%
  filter(surface %in% c("bench", "entrance_door", "handrail", "ticket_machine")) %>% 
  rownames()

adonis2(dist_subset(bray_dist, surface_subs) ~ surface,
        data = meta[surface_subs, ], permutations = 999)
```

```{r}
adonis2(dist_subset(jaccard_dist, surface_subs) ~ surface,
        data = meta[surface_subs, ], permutations = 999)
```

## Material

```{r}
material_subs <- meta %>%
  filter(surface_material %in% c("metal", "plastic", "wood")) %>% 
  rownames()

adonis2(dist_subset(bray_dist, material_subs) ~ surface_material,
        data = meta[material_subs, ], permutations = 999)
```

```{r}
adonis2(dist_subset(jaccard_dist, material_subs) ~ surface_material,
        data = meta[material_subs, ], permutations = 999)
```

# Identify the main clusters

Clustering samples on presence-absence profiles with Jaccard dist.

```{r}
bi_brackmat <- ifelse(as.matrix(abund_df) == 0, 0, 1)
samples_order <- colnames(bi_brackmat)

distm_bi <- bi_brackmat %>% t()%>%
  vegdist(method = "jaccard", diag = T, upper = T) %>% as.matrix()
colnames(distm_bi) <- samples_order
rownames(distm_bi) <- samples_order

sum(is.na(distm_bi))
class(distm_bi)
dim(distm_bi)
isSymmetric(distm_bi)
```

```{r fig.width=12, fig.height=8}
hc <- hclust(as.dist(distm_bi), method = "average")
as.dendrogram(hc)

plot(hc, hang = -1)
```

```{r}
clust <- cutree(hc, k = 3)
meta$clust <- clust[as.character(meta$sl_name)]
```

```{r}
unique(ktax$taxlvl)
```

# Creating phyloseq object

```{r}
# set station as reference for station_specificcar
#meta$station_specific <- factor(meta$station_specific)
#meta$station_specific <- relevel(meta$station_specific, ref = "car")

abund <- otu_table(as.matrix(abund_df), taxa_are_rows=T)
meta <- sample_data(meta)
tax <- tax_table(as.matrix(ktax))
#tre1 <- read_tree(my_tree_file)

pseq <- phyloseq(abund, meta, tax)
```

```{r}
all(sample_names(pseq) %in% rownames(meta))
all(rownames(meta) %in% sample_names(pseq)) 
table(meta$station) 
```

```{r}
saveRDS(pseq, file = "out/DA_analysis/pseq.rds")
```

# Run ANCOMBC

```{r}
library(openxlsx)
wb <- createWorkbook()
```


## DA stations

```{r}
pseq <- readRDS(file = "out/DA_analysis/pseq.rds")
```

```{r}
out <- ancombc2(pseq, fix_formula = "station", group = "station",
                #tax_level = "Species",
                struc_zero = TRUE,
                neg_lb = FALSE,
                p_adj_method = "holm",
                verbose = TRUE,
                pseudo_sens = TRUE,
                global = TRUE, pairwise = TRUE, dunnet = TRUE)
```

```{r}
#saveRDS(out, file = "out/DA_analysis/ANCOMBC2.stations-res.rds")

out <- readRDS(file = "out/DA_analysis/ANCOMBC2.stations-res.rds")
```

```{r save global}
addWorksheet(wb, "stations_global")

out$res_global %>% filter(diff_abn) %>% 
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T) %>%
  select("Family", "Species", "q_val", "diff_abn" , "passed_ss") %>% 
  mutate_if(is.numeric, ~formatC(., digits = 1, format = "e")) %>% 
  rename("Родина" = Family,
         "Вид" = Species,
         "P-значення" = q_val,
         "Диф.розповсюджений" = diff_abn,
         "Пройшов SS" = passed_ss) %>%
  writeData(wb, sheet = "stations_global", x = .)

saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

```{r}
res_pair <- 
out$res_pair %>%
  filter(if_any(starts_with("diff_"), ~ .x == TRUE)) %>%
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T)
```

```{r save red_pair}
addWorksheet(wb, "stations_pair")
writeData(wb, sheet = "stations_pair", x = res_pair)
saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

```{r}
heatmaps <- list()

for (i in seq_along(stations)) {
  station <- stations[i]

  lfc_cols <- res_pair %>% 
    select(starts_with("lfc_")) %>%
    select(contains(station)) %>%
    colnames()

  diff_cols <- gsub("lfc_", "diff_", lfc_cols)

  hmcols <- lfc_cols %>%
    str_remove("^lfc_") %>%
    str_remove(paste0("_?station", station, "_?")) %>% 
    str_remove("station")

  hmcols <- station_names[hmcols]
  hmcols[is.na(hmcols)] <- ref_station

  lfc_mat <- res_pair %>%
    select(Species, all_of(lfc_cols)) %>%
    column_to_rownames("Species") %>%
    as.matrix()

  diff_mat <- res_pair %>%
    select(Species, all_of(diff_cols)) %>%
    column_to_rownames("Species") %>%
    as.matrix()

  colnames(lfc_mat) <- hmcols
  colnames(diff_mat) <- hmcols
  lfc_mat <- lfc_mat[rownames(diff_mat), ]

  hmap <- Heatmap(lfc_mat,
                  name = paste0("LFC_", station),  # Unique name
                  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
                  cluster_columns = FALSE,
                  cluster_rows = FALSE,
                  show_column_names = TRUE,
                  show_row_names = (i == 1),  # Only first heatmap shows row names
                  column_title = station_names[station],
                  column_names_rot = 15,
                  row_names_side = "left",
                  cell_fun = function(j, i, x, y, w, h, fill) {
                    if (!is.na(diff_mat[i, j]) && diff_mat[i, j]) {
                      grid.text("*", x, y, gp = gpar(fontsize = 14))
                    }
                  })
  
  heatmaps[[station]] <- hmap
}

# Combine all
draw(Reduce(`+`, heatmaps), heatmap_legend_side = "right", merge_legend = TRUE)

```

```{r}
# * * * Create columns * * *
stations_ukr = c("kontraktova_ploshcha" = "Контрактова",
                 "petrivka" = "Почайна",
                 "heroiv_dnipra" = "Гер.Дніпра",
                 "minska" = "Мінська",
                 "maidan_nezalezhnosti" = "Майдан")

lfc_cols <- res_pair %>% select(starts_with("lfc_")) %>% colnames()

diff_cols <- gsub("lfc_", "diff_", lfc_cols)
ss_cols <- gsub("lfc_", "passed_ss_", lfc_cols)

hm_cols <- lfc_cols %>% 
  str_remove("lfc_station") %>% 
  str_replace("_station", " - ") %>% 
  if_else(!grepl(" - ", .), paste0(., " - heroiv_dnipra"), .)

hm_cols <- sapply(hm_cols, USE.NAMES = F, function(x) str_replace_all(x, stations_ukr))

# * * * Create matrices * * *
lfc_mat <- res_pair %>%
  select(Species, all_of(lfc_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

diff_mat <- res_pair %>%
  select(Species, all_of(diff_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

ss_mat <- res_pair %>%
  select(Species, all_of(ss_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

colnames(lfc_mat) <- hm_cols
colnames(diff_mat) <- hm_cols
colnames(ss_mat) <- hm_cols

lfc_mat <- lfc_mat[rownames(diff_mat), ]
ss_mat <- ss_mat[rownames(ss_mat), ]
```

```{r fig.width = 10, fig.height = 5}

# * * * Plot the heatmap * * *
cairo_pdf(file = "figures/ancombc2.stations.pairwise-test.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 4.5) # The height of the plot in inches

Heatmap(lfc_mat,
                name = "LFC",
                col = colorRamp2(c(-3, 0, 3), c("#26576f", "white", "#8a2f2f")),
                cluster_columns = FALSE, cluster_rows = FALSE,
                show_column_names = TRUE,
                show_row_names = TRUE,
                column_title = "ANCOMBC2: Pairwise тест між станціями",
                column_names_rot = 20,
         width = unit(180, "mm"), height = unit(72, "mm"),
                row_names_side = "left",
                cell_fun = function(j, i, x, y, w, h, fill) {
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && !ss_mat[i, j]) {
                    grid.text("*", x, y, gp = gpar(fontsize = 10))
                  }
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && ss_mat[i, j]) {
                    grid.text("**", x, y, gp = gpar(fontsize = 10))
                  }
                })

dev.off()
```
```{r}
t_lfc_mat <- t(lfc_mat)
t_diff_mat <- t(diff_mat)
t_ss_mat <- t(ss_mat)
```

```{r fig.width = 5.5, fig.height=7}
cairo_pdf(file = "figures/ancombc2.stations.pairwise-test_transposed.pdf",
    width = 5.5,
    height = 7) # The height of the plot in inches

Heatmap(t_lfc_mat,
        name = "LFC",
        show_heatmap_legend = FALSE,
        column_labels = sapply(strsplit(colnames(t_lfc_mat), " "), function(x) paste0(substr(x[1], 1, 1), ". ", x[2])),
                col = colorRamp2(c(-3, 0, 3), c("#26576f", "white", "#8a2f2f")),
                cluster_columns = FALSE, cluster_rows = FALSE,
                show_column_names = TRUE,
                show_row_names = TRUE,
                #column_title = "ANCOMBC2: Pairwise тест між станціями",
                column_names_rot = 90,
         width = unit(70, "mm"), height = unit(110, "mm"),
                row_names_side = "left",
                cell_fun = function(j, i, x, y, w, h, fill) {
                  if (!is.na(t_diff_mat[i, j]) && t_diff_mat[i, j] && !t_ss_mat[i, j]) {
                    grid.text("*", x, y, gp = gpar(fontsize = 10))
                  }
                  if (!is.na(t_diff_mat[i, j]) && t_diff_mat[i, j] && t_ss_mat[i, j]) {
                    grid.text("**", x, y, gp = gpar(fontsize = 10))
                  }
                })
dev.off()
```

## DA station_specific

```{r}
pseq <- readRDS(file = "out/DA_analysis/pseq.rds")
```

```{r}
out <- ancombc2(pseq, fix_formula = "station_specific",
                #group = "station_specific",
                #tax_level = "Species",
                #struc_zero = TRUE,
                neg_lb = FALSE,
                p_adj_method = "holm",
                verbose = TRUE,
                pseudo_sens = TRUE,
                global = FALSE, pairwise = FALSE, dunnet = FALSE)
```

```{r}
#saveRDS(out, file = "out/DA_analysis/ANCOMBC2.station_specific-res.rds")
out <- readRDS("out/DA_analysis/ANCOMBC2.station_specific-res.rds")
```

```{r eval=FALSE, include=FALSE}
# reference level -- station
addWorksheet(wb, "station_specificcar")

out$res %>% 
  filter(diff_station_specificcar) %>% 
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T) %>%
  select("Family", "Species", "q_station_specificcar",
         "diff_station_specificcar" , "passed_ss_station_specificcar") %>%
  arrange(desc(passed_ss_station_specificcar)) %>%
  mutate_if(is.numeric, ~formatC(., digits = 1, format = "e")) %>% 
  rename("Родина" = Family,
         "Вид" = Species,
         "P-значення" = q_station_specificcar,
         "Диф.розповсюджений" = diff_station_specificcar,
         "Пройшов SS" = passed_ss_station_specificcar) %>% 
  writeData(wb, sheet = "station_specificcar", x = .)

saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

## DA surface_material

```{r}
pseq <- readRDS(file = "out/DA_analysis/pseq.rds")
```

```{r}
table(pseq@sam_data[["surface_material"]])
```

```{r}
pseq_subs <-
  subset_samples(pseq,surface_material %in% c("metal", "plastic", "wood"))
pseq_subs
```


```{r}
out <- ancombc2(pseq_subs,
                   fix_formula = "surface_material",
                   group = "surface_material",
                   #tax_level = "Species",
                   struc_zero = TRUE,
                   neg_lb = FALSE,
                   p_adj_method = "holm",
                   verbose = TRUE,
                   pseudo_sens = TRUE,
                   global = TRUE, pairwise = TRUE, dunnet = FALSE)
```

```{r}
#saveRDS(out, file = "out/DA_analysis/ANCOMBC2.surface_material-res.rds")

out <- readRDS(file = "out/DA_analysis/ANCOMBC2.surface_material-res.rds")
```

```{r}
wb <- loadWorkbook("out/DA_analysis/ANCOMBC_res.xlsx")
addWorksheet(wb, "surface_material_global")

out$res_global %>% filter(diff_abn) %>% 
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T) %>%
  select("Family", "Species", "q_val", "diff_abn" , "passed_ss") %>% 
  mutate_if(is.numeric, ~formatC(., digits = 1, format = "e")) %>% 
  rename("Родина" = Family,
         "Вид" = Species,
         "P-значення" = q_val,
         "Диф.розповсюджений" = diff_abn,
         "Пройшов SS" = passed_ss) %>% 
  writeData(wb, sheet = "surface_material_global", x = .)

saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

```{r}
out$res_pair %>% 
  filter(if_any(starts_with("diff_"), ~.x))
```

```{r}
res_pair <- 
out$res_pair %>%
  filter(if_any(starts_with("diff_"), ~ .x == TRUE)) %>%
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T)
```

```{r}
#addWorksheet(wb, "surface_material_pair")
res_pair %>% 
  filter(if_any(starts_with("diff_"), ~.x)) %>% 
  writeData(wb, sheet = "surface_material_pair", x = .)
saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

```{r}
# * * * Create columns * * *
materials_ukr = c("metal" = "метал",
                  "plastic" = "пластик",
                  "wood" = "дерево")

lfc_cols <- res_pair %>% select(starts_with("lfc_")) %>% colnames()

diff_cols <- gsub("lfc_", "diff_", lfc_cols)
ss_cols <- gsub("lfc_", "passed_ss_", lfc_cols)

hm_cols <- lfc_cols %>% 
  str_remove("lfc_surface_material") %>% 
  str_replace("_surface_material", " - ") %>% 
  if_else(!grepl(" - ", .), paste0(., " - metal"), .)

hm_cols <- sapply(hm_cols, USE.NAMES = F, function(x) str_replace_all(x, materials_ukr))

# * * * Create matrices * * *
lfc_mat <- res_pair %>%
  select(Species, all_of(lfc_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

diff_mat <- res_pair %>%
  select(Species, all_of(diff_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

ss_mat <- res_pair %>%
  select(Species, all_of(ss_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

colnames(lfc_mat) <- hm_cols
colnames(diff_mat) <- hm_cols
colnames(ss_mat) <- hm_cols

lfc_mat <- lfc_mat[rownames(diff_mat), ]
ss_mat <- ss_mat[rownames(ss_mat), ]
```

```{r fig.width = 6.5, fig.height = 7}

# * * * Plot the heatmap * * *
cairo_pdf(file = "figures/ancombc2.surface_material.pairwise-test.pdf",
    width = 6.5, # The width of the plot in inches
    height = 7) # The height of the plot in inches

# shorten the name of the species
row_labels <- rownames(lfc_mat)
row_labels[row_labels == "Spiroplasma endosymbiont of Diplazon laetatorius"] <- "Spiroplasma endosymbiont"

material_hm <- Heatmap(lfc_mat,
                name = "LFC",
                col = colorRamp2(c(-3, 0, 3), c("#26576f", "white", "#8a2f2f")),
                cluster_columns = FALSE, cluster_rows = FALSE,
                show_column_names = TRUE,
                show_row_names = TRUE,
        row_labels = row_labels,
                #column_title = "ANCOMBC2: Pairwise тест між матеріалами",
                column_names_rot = 10,
        #column_names_centered = TRUE,
         width = unit(70, "mm"), height = unit(160, "mm"),
                row_names_side = "left",
                cell_fun = function(j, i, x, y, w, h, fill) {
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && !ss_mat[i, j]) {
                    grid.text("*", x, y, gp = gpar(fontsize = 10))
                    } else if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && ss_mat[i, j]) {
                      grid.text("**", x, y, gp = gpar(fontsize = 10))
                      }
                })
material_hm

dev.off()
```


```{r fig.width = 6, fig.height = 7}
Heatmap(lfc_mat,
                name = "LFC",
                col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
                cluster_columns = FALSE, cluster_rows = FALSE,
                show_column_names = TRUE,
                show_row_names = TRUE,
        row_labels = row_labels,
                #column_title = "ANCOMBC2: Pairwise тест між матеріалами",
                column_names_rot = 10,
        #column_names_centered = TRUE,
         width = unit(70, "mm"), height = unit(160, "mm"),
                row_names_side = "left",
                cell_fun = function(j, i, x, y, w, h, fill) {
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && !ss_mat[i, j]) {
                    grid.text("*", x, y, gp = gpar(fontsize = 10))
                    } else if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && ss_mat[i, j]) {
                      grid.text("**", x, y, gp = gpar(fontsize = 10))
                      }
                })
```

```{r}
lfc_cols <- res_pair %>% select(starts_with("lfc_")) %>% colnames()

diff_cols <- gsub("lfc_", "diff_", lfc_cols)
ss_cols <- gsub("lfc_", "passed_ss_", lfc_cols)

hm_cols <- lfc_cols %>% 
  str_remove("lfc_station") %>% 
  str_replace("_station", " - ") %>% 
  if_else(!grepl(" - ", .), paste0(., " - heroiv_dnipra"), .)

hm_cols <- sapply(hm_cols, USE.NAMES = F, function(x) str_replace_all(x, stations_ukr))

# * * * Create matrices * * *
lfc_mat <- res_pair %>%
  select(Species, all_of(lfc_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

diff_mat <- res_pair %>%
  select(Species, all_of(diff_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

ss_mat <- res_pair %>%
  select(Species, all_of(ss_cols)) %>%
  column_to_rownames("Species") %>%
  as.matrix()

colnames(lfc_mat) <- hm_cols
colnames(diff_mat) <- hm_cols
colnames(ss_mat) <- hm_cols

lfc_mat <- lfc_mat[rownames(diff_mat), ]
ss_mat <- ss_mat[rownames(ss_mat), ]
```

```{r fig.width = 10, fig.height = 5}

# * * * Plot the heatmap * * *
cairo_pdf(file = "figures/ancombc2.stations.pairwise-test.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 4.5) # The height of the plot in inches

Heatmap(lfc_mat,
                name = "LFC",
                col = colorRamp2(c(-3, 0, 3), c("#26576f", "white", "#8a2f2f")),
                cluster_columns = FALSE, cluster_rows = FALSE,
                show_column_names = TRUE,
                show_row_names = TRUE,
                column_title = "ANCOMBC2: Pairwise тест між станціями",
                column_names_rot = 20,
         width = unit(180, "mm"), height = unit(72, "mm"),
                row_names_side = "left",
                cell_fun = function(j, i, x, y, w, h, fill) {
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && !ss_mat[i, j]) {
                    grid.text("*", x, y, gp = gpar(fontsize = 10))
                  }
                  if (!is.na(diff_mat[i, j]) && diff_mat[i, j] && ss_mat[i, j]) {
                    grid.text("**", x, y, gp = gpar(fontsize = 10))
                  }
                })

dev.off()
```

## DA suface

```{r}
pseq <- readRDS(file = "out/DA_analysis/pseq.rds")
```

```{r}
table(pseq@sam_data[["surface"]])
```

```{r}
pseq_subs <-
  subset_samples(pseq, surface %in% c("bench", "entrance_door", "handrail", "ticket_machine"))

pseq_subs
```


```{r}
out <- ancombc2(pseq_subs,
                   fix_formula = "surface",
                   group = "surface",
                   #tax_level = "Species",
                   struc_zero = TRUE,
                   neg_lb = FALSE,
                   p_adj_method = "holm",
                   verbose = TRUE,
                   pseudo_sens = TRUE,
                   global = TRUE, pairwise = TRUE, dunnet = FALSE)
```

```{r}
saveRDS(out, file = "out/DA_analysis/ANCOMBC2.surface-res.rds")

out <- readRDS(file = "out/DA_analysis/ANCOMBC2.surface-res.rds")
```

```{r}
wb <- loadWorkbook("out/DA_analysis/ANCOMBC_res.xlsx")
addWorksheet(wb, "surface_global")

out$res_global %>% filter(diff_abn) %>% 
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T) %>%
  select("Family", "Species", "q_val", "diff_abn" , "passed_ss") %>% 
  mutate_if(is.numeric, ~formatC(., digits = 1, format = "e")) %>% 
  rename("Родина" = Family,
         "Вид" = Species,
         "P-значення" = q_val,
         "Диф.розповсюджений" = diff_abn,
         "Пройшов SS" = passed_ss) %>% 
  writeData(wb, sheet = "surface_global", x = .)

saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```

```{r}
out$res_pair %>% 
  filter(if_any(starts_with("diff_"), ~.x))
```

```{r}
res_pair <- 
out$res_pair %>%
  filter(if_any(starts_with("diff_"), ~ .x == TRUE)) %>%
  merge(ktax, by.x = "taxon", by.y = 0, all.x = T)

addWorksheet(wb, "surface_pair")
writeData(wb, sheet = "surface_pair", x = res_pair)
saveWorkbook(wb, "out/DA_analysis/ANCOMBC_res.xlsx", overwrite = T)
```


## Combined paired-test plot: material + station

```{r}
library(patchwork)

station_hm + material_hm

draw(station_hm)
draw(material_hm)
```

